A Bayesian Approach to Detecting Bias in Peremptory Challenges Using Historical Strike Data

### 2021-09-14 analysis/paper_draft_TEM.Rmd

---
title: "Detecting Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
bibliography: references.bib
author:
  - name: "Sachin S. Pandya"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0001-7387-1307
  - name: "Xiaomeng Li"
    affiliation: "University of Connecticut | Ernest & Young"
  - name: "Timothy E. Moore"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0002-9576-0517
output: distill::distill_article
---

@@

This paper presents a way to use historical data on an attorney's use of peremptory strikes in past trials to estimate, in a current trial, how likely an attorney has used peremptory challenges because of prospective juror race or gender in violation of _Batson v. Kentucky_ (1986), and its progeny.

In the US, individuals selected for jury-service appear in court as scheduled and are questioned by the parties' attorneys and the trial judge.  During this process, a prospective juror, if not excused by the trial judge for cause, may still be dismissed (declared ineligible for a seat on the jury) if a party's attorney uses a peremptory challenge against them.

# Background and Related Literature

With a peremptory challenge, a party declares a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why. While some jurisdictions have abolished peremptory strikes (Criminal Justice Act 1988, s 118(1) (UK); S.C. 2019, c. 25, s 269 (Canada)), federal and State law in the United States (except in Arizona^[Arizona Rule of Criminal Procedure XX]) authorize their use during jury selection in criminal and civil cases. In these cases, peremptory-challenge procedure can vary not only by State, but also by court and judge, including the number of peremptory challenges assigned to each side, the order of strikes, and the number of peremptory strikes each party may exercise at a time [@nscs-jury2018; @Williams2017].

Under the "struck jury" method of peremptory challenges, the trial judge decides whether to accept or reject all challenges for cause - for which a reason must be given -- before the parties exercise any peremptories. Thereafter, the number of potential jurors who remain is typically the sum of number of seats on the jury (plus alternates, if any) and the number of peremptory challenges allotted to both sides.

Then, the parties exercise their strikes on those potential jurors. In one procedure, the parties submit their strikes simultaneously. In another variant, the parties strike in an alternating sequence, with the prosecution striking first. Where the number of strikes are not the same for the parties, an alternating sequence may result in one party exercising multiple strikes consecutively. Once the parties have used (or waive) their strikes, the jurors are seated based on the order in which they were selected or at random.

Under a "strike-and-replace" or "jury box" method, enough potential jurors are seated to fill the jury box. Some of them are excused for cause, and the parties decide whether or not to strike any of the remaining jurors with their peremptory challenges.  Once that is completed, new potential jurors take the seats of the excused or stricken jurors, and the process repeats until all the jury seats are filled.  At any point, strikes can be used on anyone in the jury box.

<!--

Note: Constitutionality of Canada's abolition of strikes upheld in https://decisions.scc-csc.ca/scc-csc/scc-csc/en/item/18932/index.do

Courts also vary as to the number of peremptory strikes a party exercises at a time [@Williams2017, p. 297-298]. This particularly matters where each side does not have the same number of strikes.  For example, if the prosecution has six strikes and the defendant has ten, the custom may be to allow one prosecutor strike, then _two_ defense strikes, then the second prosecutor strike, then two more defense strikes, and so on, until all the strikes are exercised.

-->

This paper focuses on _Batson_ challenges to a party's use of peremptory challenges. Under the US Supreme Court's decision in _Batson v. Kentucky_ (1986), a party's exercise of a peremptory challenge against one or more prospective jurors violates the Equal Protection Clause of the US Constitution if motivated by such jurors' race, ethnicity, gender, or certain other protected characteristics.

For a _Batson_ challenge, the challenging party must typically establish a "prima facie case", that is, enough evidence to infer that a party struck a juror because of such juror's race, gender, or other protected status. If so, the burden shifts to the challenged party to produce one or more nondiscriminatory reasons for that strike. Thereafter, the challenger must persuade the trial judge that prospective jurors' race or gender, not the proffered nondiscriminatory reason(s), was more likely than not the real motivation for the challenged strike(s) [@LaFave2020, sec. 22.3]. For similar challenges under State law, some States have reduced or eliminated the challenger's burden to establish a prima facie case [e.g., State v. Holloway, 209 Conn. 636, 646 & n.4 (1989); State v. Chapman, 317 S.C. 302, 305-06 (1995)], while other States have shifted the inquiry away from the striking party's motives to whether a hypothetical "objective observer" would find that race or ethnicity was a "factor" in that party's use of strikes [e.g., Wash. R. 37(e)].

Despite _Batson_ and its progeny, "prosecutors in criminal cases, and defense attorneys in civil case are more likely to use their challenges to dismiss African American jurors, whereas criminal defense attorneys and civil plaintiff's attorneys are more likely to exercise their strikes on white prospective jurors." [@Diamond2018]. Possible explanations include attorney fear that even raising a _Batson_ challenge raises with it the embarrassment and discomfort associated with accusing another attorney of being racist [@OBrien2019, 29; @Offit2021]. Similarly, a trial judge deciding a _Batson_ challenge may similarly be reluctant to find against opposing counsel, particularly if that prosecutor repeatedly appears before that judge.

Prior work on statistical methods for _Batson_ challenges, however, has largely focused on strike data from a single case [e.g. @Barrett2007; @Gastwirth2016; @Kadane2018; @Kadane2018a], perhaps because strike data from past cases is hard to collect. Jury-selection records are often not readily available, because they are kept, if at all, by court staff somewhere other than the case file.  Such records are also largely on paper, not electronic or machine-readable formats [@Grosso2017; @Wright2018], and otherwise requires painstaking effort to collect [e.g., @Craft2018].

Moreover, strike data on past trials may be missing information not at random [e.g., @Baldus2001, 132-134]. Possible reasons include that some jurisdictions do not require recording relevant or detailed enough information about jurors who were or could have been struck (e.g., their race or gender), thus forcing lawyers and researchers to turn to other data sources (e.g., voter registration rolls), if any, to find that information. Finally, judges and court staff vary in how often they fully complete or maintain the information they are required to record.

When collecting strike data in multiple trials, prior studies have aggregated strikes across those trials, reported the observed difference in strike rates by race or gender, and tested for the probability of observing a non-zero difference in strike rates by race or gender, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference. For example, Eisenberg [-@Eisenberg2017, 334-339] used a 2-tail Fisher's exact test on the observed difference in peremptory strikes by race and gender in a non-random sample of thirty-five trials from 1997-2012 that resulted in death sentences (out of 63 such trials). @Gastwirth2013 has criticized the use of a paired t-test to compare the differences in strike rates of Black and non-Black venire members in the aggregate. Rather, both they and @Gastwirth2014 argued for applying the Cochran–Mantel–Haenszel test by taking the difference of the sum of the observed numbers of Black venire members challenged in each trial and the corresponding sum of the expected numbers of Blacks challenged, and then comparing it to the standard deviation of the difference.

Prior studies have also modeled how much a prospective juror's race affected the odds of being struck [e.g., @Baldus2001]. For exmpale, using jury selection data for 1992-2017 for Mississippi's Fifth Circuit Court District, Craft [-@Craft2018a, 9] applied a logistic regression model to analyze how much a prospective juror's race affected the odds of being struck by the state. For studies applying other modeling techniques to this Mississippi data, see @DeCamp2020 (propensity score matching) and @Dunn2021 (optimal feature selection, optimal classification trees).

Here, we use a Bayesian approach for detecting illegal bias for _Batson_ challenges based solely on an attorney's pattern of peremptory challenges in both a current trial and that attorney's strikes in past cases. First, we specify a model of the peremptory-strike process in the court of interest. Second, we assign an initial prior distribution for a bias parameter $b$. Third, we use an attorney's strike data from past trials, as  generated by the same peremptory-strike process, to estimate a posterior distribution for $b$.  Finally, we use that posterior distribution as the prior for $b$ in the present case and update the posterior based on data on the target attorney's strikes in that present case.

In so doing, we make two main contributions. First, we use the power prior to control how much the strike history from past cases affects the posterior distribution of the bias parameter $b$. The power prior is a way to create informative prior distributions based on historical data so as to control how much historical data affects the posterior distribution of some parameter of interest. The key insight is to raise the likelihood function based on the historical data to a power parameter [@Ibrahim2000; @Ibrahim2015].

Second, we implement the analysis as prototype software that an attorney could use in real time during jury selection to assess how likely an attorney striking prospective jurors in violation of _Batson_.

# Model 1: Alternating Strikes

Following @Kadane2018 and @Barrett2007, we model a struck-jury peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence.

For any given case $i$ in which jury selection occurs, let $j$ denote a peremptory strike used, and let $\delta_{ij}$ denote whether or not a party used that strike on a person who belongs to a "cognizable class". If "race" is the bias type of interest, the cognizable class is racial minority jurors ($\delta_{ij} =1$, 0 for White jurors).  If "gender" is the bias of interest, the cognizable class is female jurors ($\delta_{ij} = 1$, 0 for male jurors).

In turn, let $c_{ij}$ denote the number of cognizable class members subject to strike; and $m_{ij}$ denote the number of cognizable class non-members subject to strike, such that $c_{ij}+m_{ij}$ is the total number of jurors potentially subject to strike. If there is no bias, the probability is $\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking a cognizable class member, and $1-\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking someone who does not belong to that class.

By adding one parameter $w$, we can measure the bias by different values of $w$ by defining the probability of a cognizable class member being struck to be $\frac{wc_{ij}}{wc_{ij}+m_{ij}}$. To avoid making the weight of the non-cognizable class be the reciprocal of the weight of cognizable class, let $b = log(w)$. Accordingly, for any given value of the bias parameter $b$, the probability of strike of a member from either class, or $Pr(\delta_{ij})$, is shown in equation \@ref(eq:model1)

@@

\begin{equation}  
L(b|\delta) = \prod_{i=1}^{n_i} \prod_{j=1}^{n_j}(\frac{(e^b)c_{ij}}{(e^b)c_{ij}+m_{ij}})^{\delta_{ij}} (\frac{m_{ij}}{(e^b)c_{ij}+m_{ij}})^{1-\delta_{ij}} (\#eq:model2)
\end{equation}

@@

Next, we assign an initial prior:

\[b \sim \text{Normal}(0,2)\]

Unlike @Kadane2018, who assigned $b$ a prior of $Uniform (-6,6)$, because the law assigning the burden of proof for _Batson_ challenges entails a weakly-informative prior.

First, for our initial prior of $N(0,2)$, we let $\mu = 0$, because the law assigns the burden of proof in a _Batson_ challenge to the party bringing the challenge as well as the initial burden of producing some relevant evidence of illegal bias (the _Batson_ prima facie case). Thus, if that challenging party in fact produces no relevant evidence of illegal bias, the law requires a trial judge to reject the challenge as unproven. This is tantamount to treating zero as the most-likely value of the bias parameter, absent any data.

Second, we take that law as requiring, absent any data, the assumption that higher degrees of illegal bias (preference) are less likely than lower degrees of bias (preference). By using a normal (Gausian) distribution with $\sigma = 2$, we encode this assumption by making values of the bias parameter less likely as they move further from zero and by making very unlikely values any value of $b$ greater than two standard deviations from the mean.

Finally, we want to incorporate historical data on strikes in past cases and adjust the weight of that historical information on the posterior distribution of the bias parameter.  Accordingly, we introduce the power prior \@ref(eq:model3)

@@

@@

The posterior distribution does not have a closed form.  Accordingly, we used the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. To measure convergence, we used traceplots, the Gelman-Rubin convergence diagnostic [@Gelman1992; @Gelman1998] and Geweke’s diagnostics [@Gewe1992].

# Simulation

@@

We also generated historical strike data of three sizes (same, double, and triple the size of current data, i.e., data on one previous trial, two previous trials) using seven different values of the bias parameter for generating that historical data ( _historical_ $b = {-3,-2,-1,0,1,2,3}$).

We let historical $b$ differ from current $b$ (our parameter of interest) to simulate noise associated with historical strike data that is incomplete not at random. Alternatively, this difference simulates a real difference in the bias parameter that depends on a feature that the current trial has but that all the past trials in the historical data do not. To illustrate, assume a prosecutor with two true bias values: $b > 0$ if the criminal defendant is Black, $b = 0$ if not. If our historical strike data contains only in cases with non-Black defendants, accounting for that historical data may lead us to underestimate prosecutor's bias $b$ in the current case with a Black defendant.

By using the power prior, we can control the weight that we put on historical data by modifying the weight parameter $a$. If $a = 1$, the historical strike data is equally weighted with the present data on strikes. If $a < 1$, the historical data matters proportionally less. We can assign $a < 1$ to evaluate how sensitive the posterior of the bias parameter is to the historical data on strikes. Accordingly, in the simulation study, we tried different values of the weight parameter $a = (0.1,0.3,0.5,0.7,1)$ to show how that parameter affected modeling accuracy.

We considered 735 scenarios in total, i.e., the combination of 7 different bias parameters of current data and of historical data, 3 different amounts of historical data and 5 different values of the power prior weight parameter $a$. For each scenario, we generated 1000 datasets and fit the model on those datasets. We recorded the mean value of the posterior mean and credible intervals of one thousand model fits for each scenario.

In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times those intervals actually contained the true value of the (current) bias parameter $b$. For instance, a 90\% coverage rate means that in 900 of 1000 model fits, a given credible interval contained what we fixed as the value of current $b$. Put another way, if we set $b>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero among the 1000 model fits.

@@

```{r b80, echo=FALSE, fig.cap="80% credible interval"}

library(ggplot2)

# INSERT simulation and plot code

load_path <- here::here("sim_resultc.csv")

sims <- readr::read_csv(load_path)
sims$a.f <- as.factor(sims$a)

sims$sig <- ifelse(sims$L80 <0 & sims$U80 >0 , "ns", 1) 
sims$sig <- ifelse(sims$L80 >0 & sims$U80 >0 , 
                                    "pos", sims$sig) 
sims$sig <- ifelse(sims$L80 <0 & sims$U80 <0 , 
                                    "neg", sims$sig) 
sims$sig.n <- 0
sims$sig.n <- ifelse(sims$sig=="pos", 1, sims$sig.n)
sims$sig.n <- ifelse(sims$sig=="neg", 1, sims$sig.n)

p = ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  # scale_fill_gradient2(expression(hat(d)), 
  #                      low = "darkred", mid = "white", 
  #                      high = "blue", 
  #                      breaks = c(-3, -2, -1, 0, 1, 2, 3), 
  #                      limits = c(-3.2, 3.2),
  #         labels = c("Neg bias", "", "", "NS","", "", "Bias"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal()+
  theme_minimal()+
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        #axis.title = element_text(size = 16), 
        #legend.title.align = 0.05, 
        #legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
p

```{r b90, echo=FALSE, fig.cap="90% credible interval"}

# INSERT simulation and plot code
sims$a.f <- as.factor(sims$a)
sims$sig <- ifelse(sims$L90 <0 & sims$U90 >0 , "ns", 1) 
sims$sig <- ifelse(sims$L90 >0 & sims$U90 >0 , 
                                    "pos", sims$sig) 
sims$sig <- ifelse(sims$L90 <0 & sims$U90 <0 , 
                                    "neg", sims$sig) 
sims$sig.n <- 0
sims$sig.n <- ifelse(sims$sig=="pos", 1, sims$sig.n)
sims$sig.n <- ifelse(sims$sig=="neg", 1, sims$sig.n)

p = ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  # scale_fill_gradient2(expression(hat(d)), 
  #                      low = "darkred", mid = "white", 
  #                      high = "blue", 
  #                      breaks = c(-3, -2, -1, 0, 1, 2, 3), 
  #                      limits = c(-3.2, 3.2),
  #         labels = c("Neg bias", "", "", "NS","", "", "Bias"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal()+
  theme_minimal()+
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        axis.title = element_text(size = 16),
        legend.title.align = 0.05, 
        legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
p
```

```{r b95,  echo=FALSE,fig.cap="95% credible interval"}

# INSERT simulation and plot code
sims$a.f <- as.factor(sims$a)
sims$sig <- ifelse(sims$L95 <0 & sims$U95 >0 , "ns", 1) 
sims$sig <- ifelse(sims$L95 >0 & sims$U95 >0 , 
                                    "pos", sims$sig) 
sims$sig <- ifelse(sims$L95 <0 & sims$U95 <0 , 
                                    "neg", sims$sig) 
sims$sig.n <- 0
sims$sig.n <- ifelse(sims$sig=="pos", 1, sims$sig.n)
sims$sig.n <- ifelse(sims$sig=="neg", 1, sims$sig.n)

p = ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  # scale_fill_gradient2(expression(hat(d)), 
  #                      low = "darkred", mid = "white", 
  #                      high = "blue", 
  #                      breaks = c(-3, -2, -1, 0, 1, 2, 3), 
  #                      limits = c(-3.2, 3.2),
  #         labels = c("Neg bias", "", "", "NS","", "", "Bias"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal()+
  theme_classic()+  
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        axis.title = element_text(size = 16), 
        legend.title.align = 0.05, 
        legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
p
```

@@

These plots indicate that our model can detect strong bias and bias ($b = \{-3, -2, 2,3\}$). When bias is moderate ($b = \{1, -1\}$), the model can detect it if the historical data has a bias parameter in the same direction. If the bias parameter of historical data is in an opposite direction, the model may provide a wrong estimate of current bias parameter.

# Model 2: Simultaneous Strikes

In some courts, both parties simultaneously exercise their peremptory challenges on the prospective jurors in a venire.  This may result in both sides using a strike each on the same potential juror.

For any given case $i$ in which jury selection occurs and a party uses $s$ number of strikes under this process, @Barrett2007 models the probability of a party striking at least $x$ cognizable-class members at random as following a hypergeometric distribution:

>>>>

---
title: "A Bayesian Approach to Detecting Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
author:
  - name: "Sachin S. Pandya"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0001-7387-1307
  - name: "Xiaomeng Li"
    affiliation: "University of Connecticut"
  - name: "Timothy E. Moore"
    affiliation: "University of Connecticut"
    orcid_id: 0000-0002-9576-0517
output: distill::distill_article
---

# Abstract
The _Batson_ decision, and subsequent decisions, forbids the use of peremptory challenges for jury selection in ways that discriminate against cognizable classes, e.g., by race, sex, ethnicity, etc. Here, we propose a Bayesian approach for detecting illegal bias for _Batson_ challenges based on an attorney's pattern of peremptory challenges in both a current trial, as well as an attorney's strikes in past cases. Specifically, we presents the use of historical data on an attorney's use of peremptory strikes in past trials to estimate, in a current trial, how likely [should this rather be based on probability?] an attorney has used peremptory challenges because of prospective juror race or gender in violation of _Batson v. Kentucky_ (1986), and its progeny. We also introduce the use of a _power prior_ to provide flexibility in the use of historical trial data. We demonstrate the utility of our model on simulated data. Finally, we present a user-friendly R Shiny application that can be used to examine jury selection data in real time. We demonstrate the app using trail data collected from courts in Connecticut.

@@

## Jury Selection Procedures

In the US, individuals selected for jury-service appear in court as scheduled and are questioned by the parties' attorneys and the trial judge. During this "voir dire" process, a prospective juror, if not excused by the trial judge for cause, may still be dismissed (declared ineligible for a seat on the jury) if a party's attorney uses a peremptory challenge against them. Under the US Supreme Court's decision in _Batson v. Kentucky_ (1986), a party's exercise of a peremptory challenge against one or more prosepctive jurors violates the Equal Protection Clause of the US Constitution if motivated by such jurors' race, gender, or certain other protected characteristics. In bringing a _Bateson_ challenge the burden is on the attorney bringing the challenge to provide evidence of bias against prospective jurors in a cognizable class.

This paper focuses on _Batson_ challenges to a party's use of peremptory challenge. With a peremptory challenge, a party declares a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why. While countries such as the United Kingdom and Canada have abolished peremptory strikes (Criminal Justice Act 1988, s 118(1); S.C. 2019, c. 25, s 269), federal and State law in the United States continue to authorize their use during jury selection in criminal and civil cases.[ repetitive] In the US, peremptory-challenge procedure can vary not only by State, but also by court and judge, including the number of peremptory challenges assigned to each side, the order of strikes, and the number of peremptory strikes a party may exercise at a time [@nscs-jury2018; @Williams2017].

Under the "struck jury" method of peremptory challenges, the trial judge rules on all challenges for cause before the parties exercise any peremptories, so that the number of potential jurors who remain is the sum of number of seats on the jury (plus alternates, if any) and the number of peremptory challenges allotted to both sides. Then, the parties exercise their strikes on those potential jurors. In one procedure, the parties submit their strikes simultaneously. In another variant, the parties strike in an alternating sequence, with the prosecution striking first. Where the number of strikes are not the same for the parties, an alternating sequence may result in one party exercising multiple strikes per turn. If a party decides not to use all its strikes, the jurors are seated based on the order in which they were selected or at random. Under a "strike-and-replace" or "jury box" method, enough potential jurors are seated to fill the jury box. Some of them are excused for cause, and the parties decide whether or not to strike any of the remaining jurors with their peremptory challenges.  Once that is completed, new potential jurors take the seats of the excused or stricken jurors, and the process repeats until all the jury seats are filled.  At any point, strikes can be used on anyone in the jury box.

<!--
Note: Constitutionality of Canada's abolition of strikes upheld in https://decisions.scc-csc.ca/scc-csc/scc-csc/en/item/18932/index.do
Courts also vary as to the number of peremptory strikes a party exercises at a time [@Williams2017, p. 297-298]. This particularly matters where each side does not have the same number of strikes.  For example, if the prosecution has six strikes and the defendant has ten, the custom may be to allow one prosecutor strike, then _two_ defense strikes, then the second prosecutor strike, then two more defense strikes, and so on, until all the strikes are exercised.
-->

## Statistical Procedure

Following @Kadane2018 and @Barrett2007, we model a struck-jury peremptory-challenge process in which each party strikes prospective jurors in an alternating sequence. For any given case $i$ in which jury selection occurs, let $j$ denote a peremptory strike used, and let $\delta_{ij}$ denote whether or not a party used that strike on a person who belongs to a "cognizable class". If "race" is the bias type of interest, the cognizable class is racial minority jurors ($\delta_{ij} =1$, 0 for White jurors).  If "gender" is the bias of interest, the cognizable class is female jurors ($\delta_{ij} = 1$, 0 for male jurors). In turn, let $c_{ij}$ denote the number of cognizable class members subject to strike; and $m_{ij}$ denote the number of cognizable class non-members subject to strike, such that $c_{ij}+m_{ij}$ is the total number of jurors potentially subject to strike. If there is no bias, the probability is $\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking a cognizable class member, and $1-\frac{c_{ij}}{c_{ij}+m_{ij}}$ for striking someone who does not belong to that class.
By adding one parameter $w$, we can measure the _bias_ by different values of $w$ by defining the probability of a cognizable class member being struck to be $\frac{wc_{ij}}{wc_{ij}+m_{ij}}$. To avoid making the weight of the non-cognizable class be the reciprocal of the weight of cognizable class, let $b = log(w)$. Accordingly, for any given value of the bias parameter $b$, the probability of strike of a member from either class, or $Pr(\delta_{ij})$, is shown in equation \@ref(eq:model1)

@@

\begin{equation}  
L(b|\delta) = \prod_{i=1}^{n_i} \prod_{j=1}^{n_j}(\frac{(e^b)c_{ij}}{(e^b)c_{ij}+m_{ij}})^{\delta_{ij}} (\frac{m_{ij}}{(e^b)c_{ij}+m_{ij}})^{1-\delta_{ij}} (\#eq:model2)
\end{equation}
<!-- \[b \sim \text{Normal}(0,2)\] -->

@@

## Incorporating prior (historical) information

In @Kadane2018, $b$ is estimated under Bayesian framework with a uniform prior from -6 to 6. In this paper, we also would like to incorporate the historical data on strikes in past cases and be able to adjust the weight of that historical information on the posterior distribution of the bias parameter.  Accordingly, we introduce the power prior \@ref(eq:model3)

@@

<!-- The prior distribution for $b$ ($\sim \text{Normal}(0,2)$] is justified on the grounds that the party bringing the Batson challenge bears the ultimate burden of persuading the trial judge of an illegal use of strikes. -->

@@

The use of the power prior allows for flexibility in how much weight historical trail information receives in estimating the bias parameter for a current trial. [list reasons why. should this be in the introduction/discussion?]

The posterior distribution does not have a closed form.  Accordingly, we used the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. To measure convergence, we used traceplots, the Gelman-Rubin convergence diagnostic [@Gelman1992; @Gelman1998} and Geweke’s diagnostics [@Gewe1992].

## Model Performance on Simulated Data

We propose the following steps to test for the presence of bias, and thus justification for bringing a _Bateson_ challenge. First, specify a model of the peremptory-strike process in the court of interest, as outlined above. Second, assign an initial prior distribution for a bias parameter $b$ (here we use a normal distribution with mean 0 and standard deviation of 2). Third, use an attorney's strike data from past trials, as  generated by the same peremptory-strike process, to estimate a posterior distribution for $b$. Finally, use that posterior distribution as the prior for $b$ in the present case and update the posterior based on data on the target attorney's strikes in that present case.

@@

We also generated historical strike data of three sizes (same, double, and triple the size of current data, i.e., data on one previous trial, two and three previous trials) using seven different values of the bias parameter for generating that historical data ( _historical_ $b = {-3,-2,-1,0,1,2,3}$).

We let historical $b$ differ from current $b$ (our parameter of interest) to simulate noise associated with historical strike data that is incomplete not at random. Alternatively, this difference simulates a real difference in the bias parameter that depends on a feature that the current trial has but that all the past trials in the historical data do not. To illustrate, assume a prosecutor with two true bias values: $b > 0$ if the criminal defendant is Black, $b = 0$ if not. If our historical strike data contains only in cases with non-Black defendants, accounting for that historical data may lead us to underestimate prosecutor's bias $b$ in the current case with a Black defendant. By using the power prior, we can control the weight that we put on historical data by modifying the weight parameter $a$. If $a = 1$, the historical strike data is equally weighted with the present data on strikes. If $a < 1$, the historical data matters proportionally less. We can assign $a < 1$ to evaluate how sensitive the posterior of the bias parameter is to the historical data on strikes. Accordingly, in the simulation study, we tried different values of the weight parameter $a = (0.1,0.3,0.5,0.7,1)$ to show how that parameter affected modeling accuracy. We considered 735 scenarios in total, i.e., the combination of 7 different bias parameters of current data and of historical data, 3 different amounts of historical data and 5 different values of the power prior weight parameter $a$. For each scenario, we generated 1000 datasets and fit the model on those datasets. We recorded the mean value of the posterior mean and credible intervals of one thousand model fits for each scenario. In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the coverage rates, i.e., the proportion of times those intervals actually contained the true value of the (current) bias parameter $b$. For instance, a 90\% coverage rate means that in 900 of 1000 model fits, a given credible interval contained what we fixed as the value of current $b$. Put another way, if we set $b>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero among the 1000 model fits.

@@

```{r b80, echo=FALSE, fig.cap="80% credible interval"}
library(ggplot2)
# INSERT simulation and plot code
load_path <- here::here("sim_resultc.csv")
sims <- readr::read_csv(load_path)
sims$a.f <- as.factor(sims$a)
sims$sig <- ifelse(sims$L80 <0 & sims$U80 >0 , "ns", 1) 
sims$sig <- ifelse(sims$L80 >0 & sims$U80 >0 , 
                                    "pos", sims$sig) 
sims$sig <- ifelse(sims$L80 <0 & sims$U80 <0 , 
                                    "neg", sims$sig) 
sims$sig.n <- 0
sims$sig.n <- ifelse(sims$sig=="pos", 1, sims$sig.n)
sims$sig.n <- ifelse(sims$sig=="neg", 1, sims$sig.n)
p = ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  # scale_fill_gradient2(expression(hat(d)), 
  #                      low = "darkred", mid = "white", 
  #                      high = "blue", 
  #                      breaks = c(-3, -2, -1, 0, 1, 2, 3), 
  #                      limits = c(-3.2, 3.2),
  #         labels = c("Neg bias", "", "", "NS","", "", "Bias"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal()+
  theme_minimal()+
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        #axis.title = element_text(size = 16), 
        #legend.title.align = 0.05, 
        #legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
p
```

```{r b90, echo=FALSE, fig.cap="90% credible interval"}
# INSERT simulation and plot code
sims$a.f <- as.factor(sims$a)
sims$sig <- ifelse(sims$L90 <0 & sims$U90 >0 , "ns", 1) 
sims$sig <- ifelse(sims$L90 >0 & sims$U90 >0 , 
                                    "pos", sims$sig) 
sims$sig <- ifelse(sims$L90 <0 & sims$U90 <0 , 
                                    "neg", sims$sig) 
sims$sig.n <- 0
sims$sig.n <- ifelse(sims$sig=="pos", 1, sims$sig.n)
sims$sig.n <- ifelse(sims$sig=="neg", 1, sims$sig.n)
p = ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  # scale_fill_gradient2(expression(hat(d)), 
  #                      low = "darkred", mid = "white", 
  #                      high = "blue", 
  #                      breaks = c(-3, -2, -1, 0, 1, 2, 3), 
  #                      limits = c(-3.2, 3.2),
  #         labels = c("Neg bias", "", "", "NS","", "", "Bias"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal()+
  theme_minimal()+
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        axis.title = element_text(size = 16),
        legend.title.align = 0.05, 
        legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
p
```

```{r b95,  echo=FALSE,fig.cap="95% credible interval"}
# INSERT simulation and plot code
sims$a.f <- as.factor(sims$a)
sims$sig <- ifelse(sims$L95 <0 & sims$U95 >0 , "ns", 1) 
sims$sig <- ifelse(sims$L95 >0 & sims$U95 >0 , 
                                    "pos", sims$sig) 
sims$sig <- ifelse(sims$L95 <0 & sims$U95 <0 , 
                                    "neg", sims$sig) 
sims$sig.n <- 0
sims$sig.n <- ifelse(sims$sig=="pos", 1, sims$sig.n)
sims$sig.n <- ifelse(sims$sig=="neg", 1, sims$sig.n)
p = ggplot()+
  geom_tile(data = sims, aes(x = d, y = d_h,
                        fill = as.factor(sig.n)),
            color="black")+
  facet_grid(h~a, labeller = label_both)+
  # scale_fill_gradient2(expression(hat(d)), 
  #                      low = "darkred", mid = "white", 
  #                      high = "blue", 
  #                      breaks = c(-3, -2, -1, 0, 1, 2, 3), 
  #                      limits = c(-3.2, 3.2),
  #         labels = c("Neg bias", "", "", "NS","", "", "Bias"))+
  scale_x_continuous(breaks = c(-3:3), expand = c(0,0))+
  scale_y_continuous(breaks = c(-3:3), expand = c(0,0))+
  xlab("Current b")+
  ylab("Historical b")+
  scale_fill_manual(values = c("white", "blue"))+
  coord_equal()+
  theme_classic()+  
  labs(fill = "coverage") +
  theme(axis.text = element_text(size =10), 
        axis.title = element_text(size = 16), 
        legend.title.align = 0.05, 
        legend.title = element_text(size = 16), 
        legend.text = element_text(size = 14))
p
```

@@

These plots indicate that our model can detect strong bias ($b = \{-3, -2, 2,3\}$). When bias is moderate ($b = \{1, -1\}$), the model can detect it if the historical data has a bias parameter in the same direction. If the bias parameter of historical data is in the opposite direction, the model may provide a wrong estimate of current bias parameter [is this true, even for the 80% CI?].

## R Shiny Application to real data

Even with agreement on the appropriate statistical analysis, it is often hard to complete the analysis in real time _during_ jury selection, unless "statisticians accompany attorneys in the jury selection process or a convenient programme or app is developed." [@Kadane2018, 12] Following this suggestion, we describe here a prototype software application ("app") that implements the approach described above using `r R.version$version.string`[@R] and shiny [@shiny]. We demonstrate the use of the app on some real data obtained from trials conducted in the state of Connecticut and compiled by the authors. For [privacy?] reasons, the attorney names have been changed, using (package name).
